---
title: "Prior Sensitivity Analysis"
date: "2026-04-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setup

```{r}
library(UsingR)
library(Bolstad)
babies_df <- babies
```

```{r}
# Remove all observations with missing or unknown values
babies_df = subset(babies_df, wt != 999 & parity != 99 & race != 99 & age != 99 &
                     ed != 9 & ht != 99 & wt1 != 999 & drace != 99 & dage != 99 &
                     ded != 9 & dht != 99 & dwt != 999 & inc != 99 & smoke != 9 &
                     time != 99 & time != 98 & number != 99 & number != 98)
```


## Model

We extend the smoking-only model to include all biologically relevant maternal
predictors: gestation period, mother's height, mother's pre-pregnancy weight,
number of cigarettes smoked per day, parity, and maternal age. Each variable
has a clear physiological motivation:

- gestation: longer pregnancies allow more time for fetal growth
- ht: taller mothers tend to have larger babies
- wt1: heavier mothers tend to have heavier babies
- number: greater cigarette consumption is associated with lower birth weight
- parity: first-born babies tend to be slightly lighter
- age: maternal age may have a modest effect on birth weight

The model is:

  wt ~ gestation + ht + wt1 + number + parity + age

There are 7 coefficients in total (including the intercept).


## Prior Sensitivity Analysis

We fit the same model under four different normal prior specifications,
holding the prior mean fixed at zero throughout. The four priors vary along
two dimensions: the scale of the variance (diffuse vs. informative) and
whether the variance is applied uniformly or asymmetrically across predictors.

Prior 1 -- Diffuse (1e4 * I): our baseline, encodes minimal prior knowledge.
Prior 2 -- Moderate (10 * I): mild regularization toward zero.
Prior 3 -- Strong (1 * I): strong shrinkage, assumes coefficients are small.
Prior 4 -- Asymmetric: diffuse on most predictors but tighter on age and
           parity, reflecting the prior belief that these variables have
           weaker and more bounded effects on birth weight than gestation
           or smoking.

```{r}
# Prior mean: no prior belief that any variable has a nonzero effect
b0_full = rep(0, 7)

# Prior 1: Diffuse (baseline)
V0_diffuse = 1e4 * diag(rep(1, 7))

# Prior 2: Moderate
V0_moderate = 10 * diag(rep(1, 7))

# Prior 3: Strong
V0_strong = 1 * diag(rep(1, 7))

# Prior 4: Asymmetric
# Intercept, gestation, ht, wt1, number: diffuse
# parity, age: tighter (prior belief that effects are modest)
V0_asym = diag(c(1e4, 1e4, 1e4, 1e4, 1e4, 1, 1))
```

```{r}
fit_diffuse  = bayes.lm(wt ~ gestation + ht + wt1 + number + parity + age,
                        data  = babies_df,
                        prior = list(b0 = b0_full, V0 = V0_diffuse))

fit_moderate = bayes.lm(wt ~ gestation + ht + wt1 + number + parity + age,
                        data  = babies_df,
                        prior = list(b0 = b0_full, V0 = V0_moderate))

fit_strong   = bayes.lm(wt ~ gestation + ht + wt1 + number + parity + age,
                        data  = babies_df,
                        prior = list(b0 = b0_full, V0 = V0_strong))

fit_asym     = bayes.lm(wt ~ gestation + ht + wt1 + number + parity + age,
                        data  = babies_df,
                        prior = list(b0 = b0_full, V0 = V0_asym))
```


## Posterior Summaries

```{r}
cat("--- Prior 1: Diffuse (1e4 * I) ---\n")
summary(fit_diffuse)

cat("--- Prior 2: Moderate (10 * I) ---\n")
summary(fit_moderate)

cat("--- Prior 3: Strong (1 * I) ---\n")
summary(fit_strong)

cat("--- Prior 4: Asymmetric ---\n")
summary(fit_asym)
```
