---
title: "Prior Sensitivity Analysis"
date: "2026-04-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r}
library(UsingR)
library(Bolstad)
babies_df <- babies
```

```{r}
# Remove all observations with missing or unknown values
babies_df = subset(babies_df, wt != 999 & parity != 99 & race != 99 & age != 99 &
                     ed != 9 & ht != 99 & wt1 != 999 & drace != 99 & dage != 99 &
                     ded != 9 & dht != 99 & dwt != 999 & inc != 99 & smoke != 9 &
                     time != 99 & time != 98 & number != 99 & number != 98)
```

## Model

We fit a multiple linear regression of birth weight (oz) on six maternal predictors:

- gestation (days): longer pregnancies produce heavier babies
- ht (inches): taller mothers tend to have larger babies
- wt1 (lbs): heavier pre-pregnancy mothers tend to have heavier babies
- number: ordinal measure of cigarettes smoked per day (0=never, ..., 8=60+)
- parity: 0 = first birth, 1 = previous birth(s)
- age (years): maternal age at delivery

```
wt ~ gestation + ht + wt1 + number + parity + age
```

There are 7 coefficients total (including intercept).

## Prior Specifications

We compare four normal prior specifications. The prior mean b0 encodes our
expectation about the direction and rough magnitude of each effect before
seeing the data. The prior variance V0 encodes our uncertainty about those
expectations. We use three prior mean vectors:

**Prior 1 -- Diffuse, zero mean (baseline)**
No prior knowledge assumed. Mean of zero on all coefficients, very large
variance. This is the same prior structure used in the smoking-only model.

**Prior 2 -- Diffuse, zero mean, moderate variance**
Same zero-mean assumption but variance scaled down to 10 * I. This provides
mild regularization without encoding any directional beliefs.

**Prior 3 -- Informative, research-backed means, diffuse variance**
Prior means are derived from published literature on each predictor's
relationship with birth weight. Variance remains diffuse so the data can
still move the posterior substantially. This prior asks: if we encode
what existing research suggests about the direction and rough magnitude
of each effect, do our posteriors change?

Prior mean derivations from literature review:
- Intercept: set to 0 (absorbed into diffuse variance)
- gestation: Oken et al. (2003) report median birth weight increments of
  ~140g per gestational week from a US national reference sample, independent
  of the CHDS. Converting to oz/day: 140g / 28.35g/oz / 7 days ~= 0.70 oz/day.
  We use 0.7. Source: Oken et al. (2003). BMC Pediatrics, 3:6. doi:10.1186/1471-2431-3-6
- ht: consistent positive association across studies; a 5cm increase in
  maternal height is associated with increases in newborn size (Polzlberger
  et al., 2016; Thame et al., 2004). We encode a weak positive prior of
  1.0 oz/inch.
- wt1: positive association well-established (Polzlberger et al., 2016).
  We encode a small positive prior of 0.05 oz/lb.
- number: OLS estimates from multiple studies find each additional cigarette
  reduces birth weight by 8-15g (Juarez & Merlo, 2013). The `number`
  variable is ordinal with each step representing ~5 more cigarettes.
  Converting: 10g x 5 / 28.35 ~= -1.76 oz per category step. We use -1.5.
  Source: Juarez S, Merlo J (2013); Perreira et al. (2017)
- parity: first births are on average 115-138g lighter than subsequent
  births (Wilcox et al., 1996; Bohn et al., 2021). Converting: 115g /
  28.35 ~= 4 oz. Since parity = 1 means prior birth (i.e. NOT first birth),
  the coefficient should be positive. We use 4.0.
  Source: Bohn et al. (2021); Magnus et al. (1994)
- age: modest positive effect up to early 30s (Stein et al., 1989).
  We encode a small positive prior of 0.1 oz/year.
  Source: Stein Z et al. (1989)

**Prior 4 -- Informative means, tighter variance on weaker predictors**
Same research-backed means as Prior 3, but variance is tightened on all of them
(variance = 10 rather than 1e4).

```{r}
# --- Prior means ---

# Prior 1 & 2: zero mean (no prior belief)
b0_zero = rep(0, 7)

# Prior 3 & 4: research-backed means
# Order: intercept, gestation, ht, wt1, number, parity, age
b0_informed = c(0,     # intercept: no prior
                0.7,   # gestation: ~0.7 oz/day (143g/week converted)
                1.0,   # ht: positive, ~1 oz/inch
                0.05,  # wt1: small positive, ~0.05 oz/lb
                -1.5,  # number: negative, ~-1.5 oz per category step
                4.0,   # parity: positive, ~4 oz for non-first birth
                0.1)   # age: small positive, ~0.1 oz/year

# --- Prior variances ---

# Prior 1: Diffuse (baseline)
V0_diffuse = 1e4 * diag(rep(1, 7))

# Prior 2: Moderate zero-mean
V0_moderate = diag(c(10, 10, 10, 10, 10, 10, 10))

# Prior 3: Diffuse variance, informed means
V0_diffuse_informed = 1e4 * diag(rep(1, 7))

# Prior 4: Informed means, tighter variance on all of them
# Order: intercept, gestation, ht, wt1, number, parity, age
V0_tight_weak = diag(c(10,   # intercept: diffuse
                        10,   # gestation: diffuse (strong signal)
                        10,   # ht: diffuse
                        10,   # wt1: diffuse
                        10,   # number: diffuse (strong signal)
                        10,     # parity: tighter (effect well-bounded ~4 oz)
                        10))    # age: tighter (effect small and well-bounded)
```

```{r}
fit1 = bayes.lm(wt ~ gestation + ht + wt1 + number + parity + age,
                data   = babies_df,
                center = FALSE,
                prior  = list(b0 = b0_zero,     V0 = V0_diffuse))

fit2 = bayes.lm(wt ~ gestation + ht + wt1 + number + parity + age,
                data   = babies_df,
                center = FALSE,
                prior  = list(b0 = b0_zero,     V0 = V0_moderate))

fit3 = bayes.lm(wt ~ gestation + ht + wt1 + number + parity + age,
                data   = babies_df,
                center = FALSE,
                prior  = list(b0 = b0_informed, V0 = V0_diffuse_informed))

fit4 = bayes.lm(wt ~ gestation + ht + wt1 + number + parity + age,
                data   = babies_df,
                center = FALSE,
                prior  = list(b0 = b0_informed, V0 = V0_tight_weak))

# bayes.lm strips names from coefficients via as.vector(); restore them
coef_names = c("(Intercept)", "gestation", "ht", "wt1", "number", "parity", "age")
names(fit1$coefficients) = coef_names
names(fit2$coefficients) = coef_names
names(fit3$coefficients) = coef_names
names(fit4$coefficients) = coef_names
```

## Posterior Summaries

```{r}
cat("=== Prior 1: Diffuse zero-mean (baseline) ===\n")
summary(fit1)

cat("\n=== Prior 2: Moderate zero-mean ===\n")
summary(fit2)

cat("\n=== Prior 3: Informed means, diffuse variance ===\n")
summary(fit3)

cat("\n=== Prior 4: Informed means, tight variance on age/parity ===\n")
summary(fit4)
```

## Comparison Table

```{r}
means1 = fit1$coefficients
means2 = fit2$coefficients
means3 = fit3$coefficients
means4 = fit4$coefficients

comparison = data.frame(
  Term          = names(means1),
  Prior_Mean    = round(b0_informed, 3),
  Post_Diffuse  = round(as.numeric(means1), 3),
  Post_Moderate = round(as.numeric(means2), 3),
  Post_Informed = round(as.numeric(means3), 3),
  Post_Tight    = round(as.numeric(means4), 3)
)

# Add a column showing direction match with prior
comparison$Dir_Match = ifelse(
  sign(comparison$Prior_Mean) == 0 | 
  sign(comparison$Post_Diffuse) == sign(comparison$Prior_Mean),
  "Yes", "No"
)

print(comparison)
```

## Visualization

```{r fig.height=6, fig.width=10}
terms = names(means1)[-1]  # exclude intercept

prior_labels = c("Diffuse\n(zero mean)",
                 "Moderate\n(zero mean)",
                 "Informed\n(diffuse var)",
                 "Informed\n(tight var)")

cols = c("#4472C4", "#ED7D31", "#A9D18E", "#FF0000")

par(mfrow = c(2, 3), mar = c(5, 4, 3, 1))

for (term in terms) {
  idx = which(names(means1) == term)
  vals = c(means1[idx], means2[idx], means3[idx], means4[idx])
  prior_val = b0_informed[idx]

  bp = barplot(
    height    = vals,
    names.arg = prior_labels,
    main      = term,
    ylab      = "Posterior Mean",
    col       = cols,
    cex.names = 0.7,
    las       = 2,
    ylim      = range(c(vals, prior_val, 0)) * 1.3
  )
  abline(h = 0, lty = 2, col = "gray40")
  # Mark the research-based prior mean as a reference line
  abline(h = prior_val, lty = 3, col = "darkgreen", lwd = 2)
}


legend("center",
       legend = c(prior_labels, "Research prior mean"),
       fill   = c(cols, NA),
       lty    = c(NA, NA, NA, NA, 3),
       lwd    = c(NA, NA, NA, NA, 2),
       col    = c(rep("black", 4), "darkgreen"),
       bty    = "n",
       cex    = 0.9)
```

## Credible Region Test

We test H0: all slope coefficients = 0 vs H1: at least one != 0.
With 6 slope coefficients, the chi-square critical value at alpha=0.025
is 14.449.

```{r}
run_cr_test = function(fit, label) {
  b1 = fit$coefficients[-1]
  V1 = fit$post.var[-1, -1]
  cr = as.numeric(t(b1) %*% solve(V1) %*% b1)
  cat(label, "-> test statistic =", round(cr, 4))
  if (cr > 14.449) {
    cat("  (reject H0)\n")
  } else {
    cat("  (fail to reject H0)\n")
  }
}

run_cr_test(fit1, "Prior 1: Diffuse zero-mean     ")
run_cr_test(fit2, "Prior 2: Moderate zero-mean    ")
run_cr_test(fit3, "Prior 3: Informed means, diffuse")
run_cr_test(fit4, "Prior 4: Informed means, tight ")
```